---
title: Practical Machine Learning Course Project Report
author: "by Anuar Imanbayev"
output:
  html_document:
    fig_height: 9
    fig_width: 9
---

## Synopsis 
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.  

In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise. 

In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants who performed barbell lifts correctly and incorrectly in 5 different ways.

## Data Preprocessing  
```{r preprocessing, cache = T}
library(randomForest)
library(corrplot)
library(caret)
library(rpart)
library(rpart.plot)
```
### Download the Data
```{r downloaddata, cache=T}
setwd("C:/Users/Anuar Imanbayev/Desktop/DataScienceCourse/DSC_Working_Directory/PML_Project")

# Training set
if(!file.exists('pml-training.csv')) {
download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
        destfile="pml-training.csv", method="libcurl")
}

# Test set
if(!file.exists('pml-testing.csv')) {
download.file(url="https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
    destfile="pml-testing.csv", method="libcurl")
}
```  
### Read the Data
After downloading the data from the data source, we read the two csv files into two data frames.  
```{r readdata, cache = T}
trainRaw <- read.csv("pml-training.csv", sep = ",")
testRaw <- read.csv("pml-testing.csv", sep = ",")
dim(trainRaw)
dim(testRaw)
```
The training data set contains 19622 observations and 160 variables.
Meanwhile, the testing data set contains 20 observations and 160 variables.
The "classe" variable in the training set is the outcome to predict. 

### Clean the data
In this step, we clean the data and get rid of observations with missing values as well as some meaningless variables.
```{r cleandata1, cache = T}
sum(complete.cases(trainRaw))
```
First, we remove columns that contain NA missing values.
```{r cleandata2, cache = T}
trainRaw <- trainRaw[, colSums(is.na(trainRaw)) == 0] 
testRaw <- testRaw[, colSums(is.na(testRaw)) == 0] 
```  
Next, we get rid of some columns that do not contribute much to the accelerometer measurements.
```{r cleandata3, cache = T}
classe <- trainRaw$classe
trainRemove <- grepl("^X|timestamp|window", names(trainRaw))
trainRaw <- trainRaw[, !trainRemove]
trainCleaned <- trainRaw[, sapply(trainRaw, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testRaw))
testRaw <- testRaw[, !testRemove]
testCleaned <- testRaw[, sapply(testRaw, is.numeric)]
```
Now, the cleaned training data set contains 19622 observations and 53 variables.
Meanwhile, the testing data set contains 20 observations and 53 variables. 
The "classe" variable is still in the cleaned training set.

### Slice the data
Then, we split the cleaned training set into a training data set (70%) and a validation data set (30%). We will use the validation data set to conduct cross validation in future steps.  
```{r slicedata, cache = T}
set.seed(16567) # For reproducibility purposes
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainData <- trainCleaned[inTrain, ]
testData <- trainCleaned[-inTrain, ]
```

## Data Modeling
We fit a predictive model for activity recognition using **Random Forest** algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general. We use **6-fold cross validation** when applying the algorithm. K at 6 is neither too large nor too small for this dataset to maintain balance in bias and variance.
```{r datamodeling1, cache = T}
controlRf <- trainControl(method="cv", 6)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=300)
modelRf
```
Then, we estimate the performance of the model on the validation data set.  
```{r datamodeling2, cache = T}
predictRf <- predict(modelRf, testData)
confusionMatrix(testData$classe, predictRf)
```
```{r datamodeling3, cache = T}
accuracy <- postResample(predictRf, testData$classe)
accuracy
oose <- 1 - as.numeric(confusionMatrix(testData$classe, predictRf)$overall[1])
oose
```
So, the estimated accuracy of the model is 99.27% and the estimated out-of-sample error is 0.73%. 

## Predicting for Test Data Set
Now, we apply the model to the original testing data set downloaded from the data source. We remove the `problem_id` column first.  
```{r testdatapredicting, cache = T}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```  

## Appendix: Figures
1. Correlation Matrix Visualization  
```{r figure1, cache = T}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```
2. Decision Tree Visualization
```{r figure2, cache = T}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot
```